Evaluating the Performance and Energy Efficiency of COSMO-ART, a Coupled Numerical Weather Forecast and Chemical Transport Model

### 2014-04-12 trunk/paper_enahpc14/Introduction.tex
adds minor change

Comprehensive  scientific assessments  of climate-warming  trends over
the  past 50  years have  shown  that the  state and  dynamics of  the
Earth's climate system have  undergone unprecedented major changes, as
evidenced  by  increases  in   global  average  atmosphere  and  ocean
temperatures,  widespread melting  of  snow and  ice,  and rising  sea
levels.   The last report  of the  Intergovernmental Panel  on Climate
Change, released  in September 2013  \citep{IPCC-2013}, concluded that
most of the observed temperature  increases since the mid 20th Century
have  been caused  by  increasing concentrations  of greenhouse  gases
resulting from  anthropogenic activities  such as fossil  fuel burning
and  deforestation.

The Earth's  surface temperature is determined by  the balance between
incoming  solar radiation and  outgoing infrared  radiation. Radiative
forcing describes any imbalance  in Earth’s radiation budget caused by
human  or natural interventions  in the  climate system.   It directly
measures the capacity of a gas  or other forcing agents to affect that
energy balance, thereby contributing to climate change.  The intensity
of  solar energy,  reflectivity  or absorption  of  gases or  surfaces
(forests, grasslands, oceans, deserts, cloud cover, ice caps, cities),
emission of heat by various  materials, all differ in their warming or
cooling contributions to climate forcing.

The radiative forcing  potential of a greenhouse gas  is determined by
its  atmospheric concentration, warming  capacity, residence  time and
spatial  distribution.   Long-lived greenhouse  gases  such as  carbon
dioxide, methane and nitrous oxide, which linger chemically stable and
homogeneously well-mixed over time scales  of a decade to thousands of
years  in the  atmosphere,  are major  human-induced climate  drivers;
whereas  short-lived gases such  as water,  suphur dioxide  and carbon
monoxide,  which are  abundant heat-trapping  gases but  physically or
chemically  reactive  to changes  in  temperature,  cause positive  or
negative  feedbacks effects which  may vary  with forcing  history and
climate  state. On a  per molecule  basis, other  gases can  have more
potent  heat-trapping ability  than carbon  dioxide  (e.g. fluorinated
gases  are 23000  times more  effective \citep{Zehner-2012}),  but are
fortunately far less abundant in the atmosphere.

One  factor   overwhelmingly  affects  the   uncertainty  in  defining
anthropogenic  radiative  forcing:  the  effects  of  aerosols  (small
airborne liquid,  solid or mixed  particles, \citealp{Putaud-2010}) in
the  atmosphere.  Although  they  are not  considered a  heat-trapping
greenhouse   gas  and   have  shorter   atmospheric   lifetimes,  they
significantly affected the atmospheric  radiative fluxes over the last
150    years   \citep{Koch-2009,   IPCC-2013}.     Enhanced   aerosols
concentrations can impact the climate system by reflecting (e.g.  pure
sulfates  and  nitrates)  or  absorbing  (e.g.   black  carbon)  solar
radiation  and  thereby exert  a  cooling  or  warming effect  on  the
Earth-atmosphere   system,  causing   a  so-called   direct  radiative
forcing     \citep{Charlson-1991,    Haywood-2000,    Ramanathan-2001,
Liao-2005,  Bangert-2012,  Lundgren-2013}.   Depending on  their  size
distribution  and chemical  composition, they  can also  act  as cloud
condensation  and   ice  nuclei,  and  thereby   influence  the  cloud
microphysical   processes  and   optical   properties  (cloud   albedo
effect,  \citealp{Twomey-1977})  through  the  contribution  to  cloud
formation.      This     results     in     changes     in     droplet
concentrations                \citep{Albrecht-1989}                and
precipitation     \citep{Rosenfeld-2000,    Khain-2008,    Khain-2009,
Pruppacher-2010,   Seifert-2012,  Tao-2012,  Lee-2012},   producing  a
so-called  negative  indirect  radiative forcing  \citep{Haywood-2000,
Lohmann-2005,  VandenHeever-2011, Rosenfeld-2013}.  Wet  scavenging of
aerosol  particles  represents  a  major  removal  mechanism  for  air
pollutants from  the atmosphere, however the  effects of anthropogenic
aerosols on  clouds and  the hydrological cycle  as well as  the cloud
lifetime    effect    are    especially    hard    to    assess    and
quantify   \citep{IPCC-2013},   and   remain   one  of   the   largest
uncertainties in climate modeling and in climate change prediction due
to  the lack  of  knowledge in  cloud feedbacks  \citep{Sherwood-2013,
Rosenfeld-2013, Lee-2013}.

Hence to  improve our understanding of  aerosol-cloud interactions and
reduce  uncertainties  of aerosol  effects  in  climate, the  research
community is making a  concerted international effort to represent the
underlying chemical  processes through models.  These  models, such as
ART (Aerosols  and Reactive Trace  gases) extension, developed  at the
Karlsruhe Institute  of Technology (KIT),  offer a key  opportunity to
reduce   the  climate  uncertainty,   particularly  on   the  regional
scale \citep{Knote-2011, Bangert-2011,  Knote-2013}.  These models can
be coupled with climate models,  such as the regional weather forecast
model COSMO (Consortium  for Small-scale Modelling), jointly developed
by  a consortium  of  European weather  centers  including the  German
weather service  DWD and MeteoSwiss,  and used in the  climate version
(COSMO-CLM)  by a  wide  research community.   The extended  COSMO-ART
model provides  a detailed description of air  pollution chemistry and
aerosol processes,  and is  mainly designed to  study air  quality and
aerosol  meteorology  feedbacks  on  short, episodic  to  annual  time
scales.  It is capable of  simulating aerosol distributions as well as
their interactions over Europe as well as other regional domains.

COSMO-ART is  computationally much more demanding than  the COSMO core
since a  large number of additional  tracers and processes  have to be
considered.  Thus this model is currently severely-limited in terms of
applicability and expensive in terms of energy consumption.  COSMO has
recently  been  ported  to  GPUs  within the  framework  of  the  High
Performance  and  High  Productivity  Computing (HP2C)  Initiative  to
optimize it  for computational and energy  efficiency.  Although these
developments will  facilitate the  application of COSMO  for numerical
weather prediction and climate  simulations, they do little to address
the  coupling with  the  ART model  extension,  for which  significant
investments  are still required  to take  it to  a similar  level. The
efficiency of ART is being  addressed in the EU Exa2Green project. The
ultimate goal  of the  project is to  deliver a prototype  code, which
provides an energy  efficiency of at least five  times of the baseline
value.   Such   an  implementation   would  allow  the   community  to
investigate critical  questions at  higher resolution and  over longer
periods, at reduced cost  to the environment.

In  Sec.~\ref{sec:1},  we  introduce  the regional  atmospheric  model
COSMO-ART  \citep{Vogel-2009}  that  accounts  for  feedbacks  between
chemistry,  aerosols, radiation, and  clouds and  additionally specify
its technical setup related to the investigated performance and energy
evaluation  methods.   Sec.~\ref{sec:2}  describes the  different  HPC
systems   and  associated  power   measurement  equipments   used  for
conducting the benchmarking study and presents energy-aware techniques
employed  to   determine  the  energy  footprint   of  the  considered
application.   Sec.~\ref{sec:3}  gathers  benchmark results  from  the
Exa2Green  project and  highlights  areas where  improvements will  be
necessary for the subsequent  baseline refactoring.  Some related work
will  be  discussed  in  Sec.~\ref{sec:4} Finally,  we  conclude  with
practical  guidelines to identifying  energy saving  opportunities and
give  directions for further  developments of  the modeling  system in
Sec.~\ref{concl}.

>>>>

While anthropogenic greenhouse gas emissions are driving unprecedented
major  climate  changes  since   the  mid  20th  Century,  one  factor
overwhelmingly  affects  the  uncertainty  in  defining  human-induced
radiative  forcing:  the  effects   of  aerosols  in  the  atmosphere.
Although they  are not considered  a heat-trapping greenhouse  gas and
have  shorter  atmospheric lifetimes,  they  significantly modify  the
global   radiation   budget   \citep{IPCC-2013}.   Enhanced   aerosols
concentrations impact the climate  system by scattering and absorption
of   solar   radiation,    thereby   exerting   a   direct   radiative
forcing  \citep{Liao-2005,  Bangert-2012,  Lundgren-2013}; or  by  the
modification of  cloud properties, cloud fraction  and surface albedo,
causing  a negative  indirect  radiative forcing  \citep{Haywood-2000,
Lohmann-2005,     VandenHeever-2011,     Rosenfeld-2013}.      Despite
considerable  progress in  global aerosol  modelling \citep{Mann-2013}
and    measurement-based    assessments   \citep{Myhre-2009},    large
uncertainties  remain  in   current  estimates  of  aerosol  radiative
forcing   \citep{Myhre-2013,    IPCC-2013,   Lee-2013,   Randles-2013,
Rosenfeld-2013, Sherwood-2013, Stier-2013}.

Hence to  improve our understanding of  aerosol-cloud interactions and
reduce  these  uncertainties,  the  research  community  is  making  a
concerted international  effort to represent  the underlying physical,
chemical  and aerosol dynamical  processes through  numerical chemical
transport  models (CTMs)  such  as ART  (Aerosols  and Reactive  Trace
gases),   developed   at  the   Karlsruhe   Institute  of   Technology
(KIT)  \citep{Vogel-2009,  Bangert-2011,  Knote-2013}.  This  regional
scale  model system  is online  coupled with  the  operational weather
forecast  model  COSMO \citep{Baldauf-2011},  jointly  developed by  a
consortium of European weather centers and used in the climate version
(COSMO-CLM) by a wide research community.

The extended  COSMO-ART model  is computationally much  more demanding
than the  COSMO core  since a large  number of additional  tracers and
processes  have  to  be  considered.   Thus this  model  is  currently
severely-limited in  terms of applicability and expensive  in terms of
energy consumption.   Although COSMO has recently been  ported to GPUs
within  the framework of  the High  Performance and  High Productivity
Computing  (HP2C)  Initiative to  optimize  it  for computational  and
energy efficiency (\url{http://www.hp2c.ch/}), significant investments
in  ART  are still  required  to  take it  to  a  similar level.   The
efficiency of COSMO-ART is being addressed in the EU Exa2Green project
(\url{http://exa2green.eu/}). The  ultimate goal of the  project is to
deliver a  prototype code, which  provides an energy efficiency  of at
least five times of the  baseline value.  Such an implementation would
allow  the  community  to  investigate critical  questions  at  higher
resolution  and   over  longer  periods,   at  reduced  cost   to  the
environment.

Some  related work  will  be discussed  in  Sec.~\ref{sec:1}. Then  in
Sec.~\ref{sec:2}, we briefly  introduce the regional atmospheric model
COSMO-ART \citep{Vogel-2009}  and specify its  technical setup related
to  the  investigated   performance  and  energy  evaluation  methods.
Sec.~\ref{sec:3} describes different power measurement equipments used
for  conducting  this  benchmarking  study and  presents  energy-aware
techniques  employed   to  determine  the  energy   footprint  of  the
considered  application.  Sec.~\ref{sec:4}  gathers  benchmark results
from  the Exa2Green  project and  highlights areas  where improvements
will be  necessary for the subsequent  baseline refactoring.  Finally,
we  conclude with  practical guidelines  to identifying  energy saving
opportunities  and give  directions  for further  developments of  the
modeling system in Sec.~\ref{concl}.
